Supplementary Materials for Minimax Optimal Regression over Sobolev Spaces via Laplacian Regularization on Neighborhood Graphs

### 2021-02-02 notes/manuscripts/AISTATS/archive/aistats_just_in_case_4.tex
Tidying up.

@@

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to non-parametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator $\wh{f}$, and a goodness-of-fit test based on $\wh{f}$. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, where $\Xset \subset \Rd$ and $1 \leq d < 4$; in the estimation problem, they are within a $\log n$ factor of optimal when $d = 4$. Additionally, we prove that Laplacian smoothing is manifold adaptive: if $\Xset$ is an $m$ dimensional manifold embedded in $\Reals^d$ with $m \ll d$, then the error rate of Laplacian smoothing methods depends only on the intrinsic dimension $m$ and not on the ambient dimension $d$. Empirically, we show that Laplacian smoothing appears not to be minimax optimal when $d > 4$. 
\end{abstract}

\section{INTRODUCTION}

Suppose we observe responses according to the signal plus noise model:
\begin{equation}
\label{eqn:signal_plus_noise_model}
Y_i = f_{0,i} + \varepsilon_i
\end{equation}
with $\varepsilon_i \sim N(0,1)$ independent Gaussian noise. Our goal is to perform statistical inference on the unknown regression function $f_0$, by which we mean either \emph{estimating} $f_0$ or more simply \emph{testing} whether $f_0 = 0$, i.e whether there is any signal present.

The Laplacian smoothing estimator $\wh{f}$ \citep{smola2003} is a penalized least squares estimator, defined on a graph. Letting $G = \bigl(\{1,\ldots,n\},W\bigr)$ be a weighted graph over the vertices $\{1,\ldots,n\}$, 
\begin{equation}
\label{eqn:laplacian_smoothing}
\wh{f} :=  \min_{f \in \Reals^n} \biggl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T \Lap f \biggr\},
\end{equation}
where $\Lap$ is the graph Laplacian matrix (defined in Section 2), and the penalty
\begin{equation*}
f^T \Lap f = \frac{1}{2} \sum_{i,j = 1}^{n} W_{ij}\bigl(Y_i - Y_j\bigr)^2
\end{equation*}
encourages $\wh{f}_i \approx \wh{f}_j$ when $W_{ij}$ is large. Assuming~\eqref{eqn:laplacian_smoothing} is a reasonable estimator of $f_0$, the squared empirical norm
\begin{equation}
\label{eqn:laplacian_smoothing_test}
\wh{T} = \frac{1}{n}\bigl\|\wh{f}\bigr\|_2^2 
\end{equation}
is in turn a reasonable statistic to assess whether or not $f_0 = 0$.

Of course there exist many methods for nonparametric regression, but Laplacian smoothing has its own set of advantages. For instance:
\begin{itemize}
	\item \emph{Computational ease.} Laplacian smoothing is fast, easy, and stable to compute. The estimate $\wh{f}$ can be computed by solving a symmetric, diagonally dominant linear system. There are known extremely fast and reliable solvers for exactly this problem (see e.g. the seminal papers of \cite{spielman2011,spielman2013,spielman2014}, or the monograph by~\cite{vishnoi2012} and references therein).
	\item \emph{Generality.} Laplacian smoothing is well defined whenever one can associate a graph with some observed responses. This generality lends itself to many different data modalities: for instance text and image classification, to name only two \citep{kondor2002, belkin03a,belkin2006} \textcolor{red}{(TODO)}: Add a bunch more references here.
	\item \emph{Weak supervision.} Although we study Laplacian smoothing in the supervised learning setup~\eqref{eqn:signal_plus_noise_model}, it can be easily adapted to the semi-supervised or unsupervised settings \citep{nadler09,slepcev17,dunlop2020,calder2019b}.
\end{itemize}

For these reasons, a body of work has emerged analyzing the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, these works can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item In the \emph{discrete (fixed design) perspective}, both $G$ and $f_0 \in \Reals^n$ are treated as fixed, and the regression ``function'' is really a vector. In this context, tight upper bounds have been derived on the error of graph-based estimators \citep{wang2016, sadhanala16,sadhanala17,kirichenko2017,kirichenko2018} and tests \citep{sharpnack2013,sharpnack2013b,sharpnack2015}, which certify that such procedures are \emph{optimal} over ``function'' classes $\mc{F} \subseteq \Reals^n$. The downside of this work is that, depending on the context, it may be unnatural to assume a priori that the graph $G$ is fixed, or that the signal $f_0$ belongs to some discrete ``function'' class.
	\item In the arguably more natural \emph{continuum (random design) perspective}, in addition to the noisy responses $Y_i$ one observes randomly sampled design points $\{X_1,\ldots,X_n\}$ belonging to a domain $\Xset \subseteq \Rd$, and assumes that $f_{0,i} \equiv f_0(X_i)$ are $n$ evaluations of a regression function $f_0: \Xset \to \Reals$. The user then builds a neighborhood graph over the design points, so that $W_{ij}$ is large iff $X_i$ and $X_j$ are close, say in Euclidean distance. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a \emph{continuum} limit (see \citep{belkin07,trillos2018,vonluxburg2008} among others). However, until recently such statements were not accompanied by error rates, and such error rates as have been proved \citep{lee2016,trillos2020} are not optimal.
\end{itemize}
Viewed as a whole, this body of work leaves a key question unanswered: what are the optimality properties of graph Laplacian smoothing, viewed as an estimator of a (continuum) regression function? This is no small point---rather, it is arguably the fundamental question of nonparametric regression, and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to its competitors.

@@

\begin{table}
\begin{center}
	\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
		Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing_test} & Thin-plate splines \eqref{eqn:thin_plate_spline} \\
		\hline
		$d = 1$ & $n^{-4/5}$ & \textcolor{red}{$n^{-4/5}$} \\
		$d = 2,3$ & $n^{-4/(4 + d)}$ & \textcolor{red}{$n^{-1/2}$} \\
		$d \geq 4$\footnotemark & $n^{-1/4}$ & \textcolor{red}{$n^{-1/4}$}
	\end{tabular}
\end{center}
\caption{Summary of testing rates over first-order Sobolev balls; rates in black are new results of this paper. When $\Xset$ is an $m$ dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$. Rates for $d \geq 4$ assume $f_0 \in \Leb^4(\Xset,M)$.}
\label{tbl:testing_rates}
\end{table}
\footnotetext{Assuming $f_0 \in \Leb^4(\Xset,M)$.}

@@

\paragraph{Sobolev spaces.}
Now that we've defined the neighborhood graph Laplacian $\Lap_{n,r}$, we step away from graph-based methods for a moment, to briefly recall some classical results regarding minimax rates over Sobolev spaces.

We say that a function $f \in \Leb^2(\Xset)$ belongs to the \emph{first-order Sobolev space} $H^1(\Xset)$ iff for each $j = 1,\ldots,d$, the weak partial derivative $D^jf$ exists and belong to $\Leb^2(\Xset)$. For such functions $f \in H^1(\Xset)$, the Sobolev seminorm $\seminorm{f}_{H^{1}(\Xset)}$ is the average size of the gradient;
\begin{equation*}
\seminorm{f}_{H^1(\Xset)}^2 = \int_{\Xset} \bigl\|\nabla f(x)\bigr\|^2 \,dx
\end{equation*}
with corresponding Sobolev norm $\norm{f}_{H^1(\Xset)} := \norm{f}_{\Leb^2(\Xset)} + |f|_{H^1(\Xset)}$. The Sobolev balls $H^1(\Xset,M)$ for $M > 0$ are then defined as
\begin{equation*}
H^1(\Xset,M) := \Bigl\{f \in H^1(\Xset): \norm{f}_{H^1(\Xset)}^2 \leq M^2\Bigr\}.
\end{equation*}
For further details regarding Sobolev spaces see e.g \citep{evans10, leoni2017}.

@@

Note that as an alternative to Laplacian smoothing --- which encourages $\wh{f}_1,\ldots,\wh{f}_n$ to satisfy a notion of smoothness over $G_{n,r}$--- we could have directly asked for an estimate $\wt{f}$ which was smooth in the Sobolev sense,
\begin{equation}
\label{eqn:thin_plate_spline}
\wt{f} := \argmin_{f} \biggl\{\sum_{i = 1}^{n} \bigl(Y_i - f(X_i)\bigr)^2 + \rho \cdot \seminorm{f}_{H^{1}(\Xset)}^2\biggr\}
\end{equation}
where the minimum is taken over all $f \in H^1(\Xset)$. Connections between the penalties in~\eqref{eqn:laplacian_smoothing} and~\eqref{eqn:thin_plate_spline} have been known for some time: \cite{bousquet03} show that for any $f \in C^2(\Xset)$, 
\begin{equation}
\label{eqn:seminorm_consistency}
\begin{aligned}
\lim \frac{1}{n^2 r^{d + 2}} f^T \Lap_{n,r} f & = \int_{\Xset} f(x) \cdot \Delta_Pf(x) p(x) \,dx \\
& = \int_{\Xset} \|\nabla f(x)\|^2 p^2(x) \,dx \\
& = |f|_{H^1(\Xset)}^2;
\end{aligned}
\end{equation}
in the above, the limit is as $n \to \infty$ and $r \to 0$, $\Delta_P: C^2(\Xset) \to \Leb^2(\Xset)$ is the Laplace-Beltrami operator
\begin{equation*}
\Delta_pf(x) := -\frac{1}{p} \dive\bigl(p^2\nabla f(x))
\end{equation*}
and the second equality follows if $f$ satisfies Dirichlet boundary conditions, by integrating by parts.

>>>>

% Make a widecheck symbol (thanks, Stack Exchange!)
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
	<5> <6> <7> <8> <9> <10>
	<10.95> <12> <14.4> <17.28> <20.74> <24.88>
	mathx10
}{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareMathAccent{\wb}{0}{mathx}{"73}

@@

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to non-parametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator $\wh{f}$, and a goodness-of-fit test based on $\wh{f}$. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, where $\Xset \subset \Rd$ and $1 \leq d < 4$; in the estimation problem, they are within a $\log n$ factor of optimal when $d = 4$. Additionally, we prove that Laplacian smoothing is manifold adaptive: if $\Xset$ is an $m$ dimensional manifold embedded in $\Reals^d$ with $m \ll d$, then the error rate of Laplacian smoothing methods depends only on the intrinsic dimension $m$ and not on the ambient dimension $d$. Empirically, we show that Laplacian smoothing appears not to be minimax optimal over $H^1(\Xset)$ when $d > 4$. 
\end{abstract}

\section{INTRODUCTION}
In the nonparametric regression problem, we observe samples $(X_1,Y_1),\ldots,(X_n,Y_n)$ according to the signal plus noise model
\begin{equation}
\label{eqn:signal_plus_noise_model}
Y_i = f_{0}(X_i) + \varepsilon_i,
\end{equation}
with $\varepsilon_i \sim N(0,1)$ independent Gaussian noise. Our goal is to perform statistical inference on the unknown regression function $f_0$, by which we mean either \emph{estimating} $f_0$ or more simply \emph{testing} whether $f_0 = 0$, i.e whether there is any signal present.

The Laplacian smoothing estimator $\wh{f}$ \citep{smola2003} is a penalized least squares estimator. It is defined with respect to the weighted graph $G = \bigl(\{1,\ldots,n\},W\bigr)$ over the vertices $\{1,\ldots,n\}$ (equivalently, over the design points $X_1,\ldots,X_n$) as
\begin{equation}
\label{eqn:laplacian_smoothing}
\wh{f} :=  \min_{f \in \Reals^n} \biggl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T \Lap f \biggr\}.
\end{equation}
Here $\Lap$ is the graph Laplacian matrix (defined in Section~\ref{sec:problem_setup_and_background}), and the penalty
\begin{equation*}
f^T \Lap f = \frac{1}{2} \sum_{i,j = 1}^{n} W_{ij}\bigl(Y_i - Y_j\bigr)^2
\end{equation*}
encourages $\wh{f}_i \approx \wh{f}_j$ when $W_{ij}$ is large. Assuming~\eqref{eqn:laplacian_smoothing} is a reasonable estimator of $f_0$, the squared empirical norm
\begin{equation}
\label{eqn:laplacian_smoothing_test}
\wh{T} := \frac{1}{n}\bigl\|\wh{f}\bigr\|_2^2 
\end{equation}
is in turn a reasonable statistic to assess whether or not $f_0 = 0$.

Of course there exist many methods for nonparametric regression \citep{gyorfi2006,wasserman2006,tsybakov2008_book}, but Laplacian smoothing has its own set of advantages. For instance:
\begin{itemize}
	\item \emph{Computational ease.} Laplacian smoothing is fast, easy, and stable to compute. The estimate $\wh{f}$ can be computed by solving a symmetric, diagonally dominant linear system. There are known extremely fast and reliable solvers for exactly this problem (see e.g. the seminal papers of \cite{spielman2011,spielman2013,spielman2014}, or the monograph by~\cite{vishnoi2012} and references therein).
	\item \emph{Generality.} Laplacian smoothing is well defined whenever one can associate a graph with some observed responses. This generality lends itself to many different data modalities: for instance text and image classification, to name only two \citep{kondor2002, belkin03a,belkin2006}. 
	\item \emph{Weak supervision.} Although we study Laplacian smoothing in the supervised learning setup~\eqref{eqn:signal_plus_noise_model}, it can be easily adapted to the semi-supervised or unsupervised settings \citep{nadler09,slepcev17,dunlop2020,calder2019b}.
\end{itemize}

For these reasons, a body of work has emerged analyzing the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, these works can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item In the \emph{discrete (fixed design) perspective}, the design points $\{X_1,\ldots,X_n\}$ and the graph $G$ are treated as fixed, and inference is carried out on the evaluations $\bigl(f_0(X_1),\ldots,f_0(X_n)\bigr) \in \Reals^n$. In this context, tight upper bounds have been derived on the error of graph-based estimators \citep{wang2016, sadhanala16,sadhanala17,kirichenko2017,kirichenko2018} and tests \citep{sharpnack2013,sharpnack2013b,sharpnack2015}, which certify that such procedures are \emph{optimal} over ``function'' classes $\mc{F} \subseteq \Reals^n$. The downside of this work is that, depending on the context, it may be unnatural to assume a priori that the graph $G$ is fixed, or that the evaluations $\bigl(f_0(X_1),\ldots,f_0(X_n)\bigr)$ belongs to some discrete ``function'' class.
	\item In the arguably more natural \emph{continuum (random design) perspective},  one assumes the design points $\{X_1,\ldots,X_n\}$ are sampled independently from some distribution $P$ supported on a domain $\Xset$. Inference is drawn on the regression function $f_0: \Xset \to \Reals$, which is assumed to be smooth in some sense, e.g. it possesses a first derivative bounded in $L^{\infty}$ (Holder) or $\Leb^2$ (Sobolev) norm.

	To conduct graph-based inference on $f_0$, the user builds a neighborhood graph over the design points, so that $W_{ij}$ is large iff $X_i$ and $X_j$ are close, say in Euclidean distance, and then solves e.g.~\eqref{eqn:laplacian_smoothing} or~\eqref{eqn:laplacian_smoothing_test}. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a \emph{continuum} limit (see \citep{belkin07,trillos2018,vonluxburg2008} among others). However, until recently such statements were not accompanied by error rates, and such error rates as have been proved \citep{lee2016,trillos2020} are not optimal over continuum function spaces, such as H\"{o}lder or Sobolev classes. 
\end{itemize}
Viewed as a whole, this body of work leaves a key question unanswered:

\begin{quote}{What are the optimality properties of graph Laplacian smoothing, viewed as an estimator of a function $f_0$ that is smooth in a continuum sense?} 
\end{quote}
On the one hand, this is no small question---rather, it is arguably the fundamental question of nonparametric regression, and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to its competitors. On the other hand, it seems difficult to answer: there is a fundamental gap between the \emph{discrete} smoothness imposed by the penalty $f^T \Lap f$ in~\eqref{eqn:laplacian_smoothing} and the \emph{continuum} smoothness assumed on $f_0$, and to obtain sharp upper bounds we needs to bridge this gap in a suitably tight sense.

\section{SUMMARY OF RESULTS}
In light of this last point, it is worth stepping back for a moment and asking: do we really need to a solve a discrete problem at all? After all, we could have instead directly leveraged the (continuum) smoothness of $f_0$ by solving the following variational problem:
\begin{equation}
\label{eqn:thin_plate_spline}
\wt{f} := \argmin_{f} \biggl\{\sum_{i = 1}^{n} \bigl(Y_i - f(X_i)\bigr)^2 + \rho \cdot \int_{\Xset} \|\nabla f(x)\|^2 \,dx \biggr\}
\end{equation}

\begin{itemize}
	\item 
	Of course, there are computational advantages to Laplacian smoothing, as mentioned previously, but let's leave those aside for the second, and ask only whether there is any statistical advantage.
	\item At first blush, the more likely answer is no. As we detail in Section~\ref{sec:minimax_optimal_laplacian_smoothing}, the Laplacian smoothing problem~\eqref{eqn:laplacian_smoothing} and variational problem~\eqref{eqn:thin_plate_spline} are in fact intimately connected; indeed~\eqref{eqn:laplacian_smoothing} can be viewed as a discrete and noisy approximation to~\eqref{eqn:thin_plate_spline}. At first blush, this suggests that Laplacian smoothing should at best of inherent the statistical properties thin-plate splines, and at worst possibly have meaningfully larger error.
	\item However the actual story is quite different: remarkably, Laplacian smoothing outperforms smoothing splines! 
	\item Laplacian smoothing also demonstrates manifold 
\end{itemize}

\paragraph{Future extensions.}

@@

\begin{table}
\begin{center}
	\begin{tabular}{p{.1\textwidth} | p{.14\textwidth} p{.12\textwidth} }
		Dimension & Laplacian smoothing \eqref{eqn:laplacian_smoothing_test} & Thin-plate splines \eqref{eqn:thin_plate_spline} \\
		\hline
		$d = 1$ & $n^{-4/5}$ & \textcolor{red}{$n^{-4/5}$} \\
		$d = 2,3$ & $n^{-4/(4 + d)}$ & \textcolor{red}{$n^{-1/2}$} \\
		$d \geq 4$\footnotemark & $n^{-1/4}$ & \textcolor{red}{$n^{-1/4}$}
	\end{tabular}
\end{center}
\caption{Summary of testing rates over first-order Sobolev balls; rates in black are new results of this paper. Rates for $d \geq 4$ assume $f_0 \in \Leb^4(\Xset,M)$. When $\Xset$ is an $m$ dimensional manifold embedded in $\Rd$, all rates hold with $d$ replaced by $m$.}
\label{tbl:testing_rates}
\end{table}
\footnotetext{Assuming $f_0 \in \Leb^4(\Xset,M)$.}

@@

@@

Note that as an alternative to Laplacian smoothing --- which encourages $\wh{f}_1,\ldots,\wh{f}_n$ to satisfy a notion of smoothness over $G_{n,r}$--- we could have directly asked for an estimate $\wt{f}$ which was smooth in the Sobolev sense,

where the minimum is taken over all $f \in H^1(\Xset)$. Connections between the penalties in~\eqref{eqn:laplacian_smoothing} and~\eqref{eqn:thin_plate_spline} have been known for some time: \cite{bousquet03} show that for any $f \in C^2(\Xset)$, 
\begin{equation}
\label{eqn:seminorm_consistency}
\begin{aligned}
\lim \frac{1}{n^2 r^{d + 2}} f^T \Lap_{n,r} f & = \int_{\Xset} f(x) \cdot \Delta_Pf(x) p(x) \,dx \\
& = \int_{\Xset} \|\nabla f(x)\|^2 p^2(x) \,dx \\
& = |f|_{H^1(\Xset)}^2;
\end{aligned}
\end{equation}
in the above, the limit is as $n \to \infty$ and $r \to 0$, $\Delta_P: C^2(\Xset) \to \Leb^2(\Xset)$ is the Laplace-Beltrami operator
\begin{equation*}
\Delta_pf(x) := -\frac{1}{p} \dive\bigl(p^2\nabla f(x))
\end{equation*}
and the second equality follows if $f$ satisfies Dirichlet boundary conditions, by integrating by parts.

### 2021-02-02 notes/manuscripts/AISTATS/archive/minimax_optimal_laplacian_smoothing3.tex
Tidying up.

\textcolor{red}{(TODO)}

>>>>

In Figure~\ref{fig:fig1}, we empirically investigate the mean squared error of the Laplacian smoothing estimator, for different values of dimension $d$. In these experiments, $X_1,\ldots,X_n$ are sampled uniformly over $\Xset = [-1,1]^d$, and the regression function $f_0(x) = \cos(m \pi x)$, where $m = 6$ for $d = 1$, $m = 2$ for $d = 2$, and $m = 1$ for $d \geq 3$. For reference, we compare Laplacian smoothing to a $k$NN estimator, which is known to achieve the minimax optimal rate $n^{-2/(2+d)}$ for all $d \geq 1$.

When $d = 4$ this continues to be the case for $k$NN regression, but not for Laplacian smoothing. In this case, the \textcolor{red}{expected slope} for Laplacian smoothing over this range of $n$---based on the upper bound $\log(n)n^{-1/4}$ derived in Theorem~\ref{thm:laplacian_smoothing_estimation1}---is $.18$.

\begin{figure}[thb]
	\includegraphics[width=.23\textwidth]{figures/cosine/mse_by_sample_size_2d.pdf}
	\includegraphics[width=.23\textwidth]{figures/cosine/mse_by_sample_size_3d.pdf}
	\includegraphics[width=.23\textwidth]{figures/cosine/mse_by_sample_size_4d.pdf}
	\includegraphics[width=.23\textwidth]{figures/cosine/mse_by_sample_size_5d.pdf}

	\caption{Comparison of Laplacian smoothing and $k$NN regression, as a function of sample size $n$. Plots are on the log-log scale, MSEs are averaged over 5 iterations, with both methods tuned for optimal MSE.  When $d = 1,2,3$, empirically both methods are achieving or exceeding the minimax rate.}
	\label{fig:fig1}
\end{figure}
